Journal and article coverage

### 2025-03-04 analysis/manuscript.tex
Continuing results: publisher analysis

This study aims to demonstrate the suitability of open scholarly data
sources for assessing the impact of transformative agreements on hybrid
open access. As shown in Figure \ref{fig:workflow}, the methodology
involved comparing hoaddata, an openly available collection of open
research information on hybrid open access, with the bibliometric
databases Web of Science and Scopus. This section introduces the initial
data sources followed by a presentation of the necessary pre-processing
steps to obtain eligible articles from transformative agreements by
using author roles (first and corresponding) and harmonised affiliation
data.

@@

As discussed elsewhere (Jahn, 2025), hoaddata uses a unified dataset of
historic JCT snapshots since July 2021, which were enriched with ISSN
variants linked to an ISSN-L and associated ROR-IDs representing
university hospitals or institutes of large research organisations,
according to OpenAlex' institution entity. A comprehensive exclusion of
fully open access journals using multiple journal lists (DOAJ, OpenAlex,
Bielefeld) was performed to identify hybrid journals. hoaddata relies on
Crossref for obtaining journal publication volume and open access status
through Creative Commons licence information relative to the published
version (``version of record''). Because of limited affiliation metadata
in Crossref (Eck \& Waltman, 2022), hoaddata sources first-author
affiliations from OpenAlex.

hoaddata follows good practices for computational reproducibility using
R (Marwick et al., 2018). The package, which includes data, code, a test
suite and documentation, is openly available on GitHub. To ensure
computational reproducibility while aggregating the data, a GitHub
Actions continuous integration and delivery (CI/CD) workflow interfaces
with the SUB Göttingen's open scholarly data warehouse based on Google
BigQuery, which provides high-performant programmatic access to monthly
snapshots of Crossref and OpenAlex. The workflow has run regularly to
fetch updates from these data sources since 2022. The package version
used in this study is 0.3, containing data from the Crossref 2024-08
dump provided to Crossref Metadata Plus subscribers and the OpenAlex
2024-08-29 monthly dump. This version including the computation log is
available on GitHub
(https://github.com/subugoe/hoaddata/releases/tag/v.0.3).

@@

Article metadata were retrieved from each database using all ISSN
variants linked to an ISSN-L. The metadata included DOIs, publication
dates, article types, open access information as well as author roles
and institutional affiliations. Publication years were determined using
the earliest known date of publication in a journal. articles in
hoaddata, this corresponded to Crossref's issued date. For Web of
Science and Scopus, the earliest publication date was used where
available, with Scopus dates specifically determined by the KB through
version tracking of the raw data.

@@

Published articles were considered hybrid open access when they were
made freely available under a Creative Commons license on publishers'
platforms. While hoaddata sourced this information from Crossref license
metadata, Web of Science and Scopus relied on Unpaywall. Unpaywall
supplements Crossref metadata by parsing publisher websites directly,
addressing cases where publishers do not provide machine-readable
Creative Commons license information (Piwowar et al., 2018). This
additional parsing remains necessary despite transformative agreement
workflows requiring license information during DOI registration
(Geschuhn \& Stone, 2017). Both Web of Science and Scopus defined hybrid
open access consistently as content available under Creative Commons
licenses on publisher platforms, distinguishing it from bronze open
access that lack such explicit license information.

@@

This process linked 6,375 ROR IDs to 4,894 Scopus Affiliation IDs, and
6,034 ROR IDs to 4,894 enhanced affiliation strings in the Web of
Science. Quality evaluation through random sampling of 50 pairs revealed
an error rate of 22\% for Web of Science (11 mismatches) and 6\% for
Scopus (3 mismatches). Upon inspection, these mismatches primarily
occurred with less-represented institutions having only a few
publications, introduced through multiple affiliations of single
authors. The difference between databases suggests that Scopus's
affiliation control aligns more closely with ROR than that of the Web of
Science.

@@

As a result of the comprehensive data processing described above,
datasets on open access in hybrid journals included in transformative
agreements were aggregated for each database at country and journal
level by year. Table~\ref{tbl-methods_overview_tab} provides a general
overview of the coverage between 2019 and 2023 per database. It shows
that the majority of hybrid journals published at least one original
research article or review marked as core during the five-year period.
These journals formed the basis for the subsequent aggregation of
publication metrics.

@@

\centering
\begin{tabular}[t]{lrrr}
\toprule
\textbf{} & \textbf{HOAD} & \textbf{Web of Science} & \textbf{Scopus}\\
\midrule
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Hybrid journal metrics}}\\
\hspace{1em}Active journals & 12,890 & 8,655 & 11,888\\
\hspace{1em}Active journals (core) & 12,888 & 8,655 & 11,878\\
\hspace{1em}Active journals (core) with OA & 11,348 & 8,392 & 11,313\\
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Publication metrics}}\\
\hspace{1em}Total published articles & 9,740,015 & 8,616,053 & 8,117,644\\
\hspace{1em}Core articles & 8,158,425 & 6,708,083 & 7,317,703\\
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Digital Object Identifier (DOI) coverage}}\\
\hspace{1em}Articles with DOI & 9,740,015 & 7,713,796 & 8,105,112\\
\hspace{1em}Core articles with DOI & 8,158,425 & 6,695,661 & 7,314,327\\
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Open Access (OA) metrics}}\\
\hspace{1em}OA articles & 998,699 & 1,112,758 & 974,099\\
\hspace{1em}Core OA articles & 969,817 & 1,019,784 & 922,578\\
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Core articles with affiliation data}}\\
\hspace{1em}First author articles & 7,242,542 & 6,294,855 & 7,232,017\\
\hspace{1em}Corresponding author articles & 5,534,207 & 6,291,441 & 6,898,487\\
\bottomrule
\end{tabular}

@@

While hoaddata only covers articles with a DOI, Scopus and Web of
Science publication metrics were aggregated using the database
identifier. A subsequent comparison of DOI coverage shows that non-core
articles in Web of Science often lacked a DOI. This was particularly the
case for meeting abstracts, which are notably prevalent in Health
Sciences journals (Melero-Fuentes et al., 2025) and are not indexed by
Scopus (Donner, 2017). Open access indicators were aggregated by DOI, as
Unpaywall only collects information on open access status for articles
with a DOI. A closer look at the core articles with affiliation data
reveals a lack of data on corresponding authors in the case of OpenAlex,
the affiliation data source used by hoaddata, compared to Web of Science
and Scopus. Only about two-thirds of the articles examined provided
corresponding author affiliations. For first authors, the proportion was
89\%. Therefore, only first author data for hoaddata were considered in
the following analysis.

@@

Figure \ref{fig-upset_coverage_results}B also shows DOIs that were only
available via hoaddata (grey area). In case of hybrid ournals covered by
all three data sources, 1,023,882 DOIs were only present in hoaddata.
After validation at the DOI level using the KB databases and manual
inspection, the main reasons for missing DOI coverage in the proprietary
database were insufficient classification of journal content as original
research articles and reviews in hoaddata. Particulalry, letters and
editorials could not be fully detected. Paratext recognition failed in
37\% of DOIs to identify non-scientific content such as front matters or
reviewer lists, which are generally not indexed by Scopus and Web of
Science. To a lesser extent, differences in publication and indexing
dates were a reason for non-overlapping DOIs.

Using overlapping DOIs, the publication volume between 2019 and 2023 was
also calculated for each journal. Figure
\ref{fig-upset_coverage_results}C shows the distribution for each
combination. It shows that there is a large spread across the journals
covered by all three data sources. Furthermore, journals in this set
published more on average than journals covered by only one or two data
sources. In particular, the journals covered exclusively by hoaddata are
substantially smaller than those covered by all three sources. Upon
inspection, these are often newly launched journals, which explains the
relatively low five-year publication volume.

@@

Figure \ref{fig-upset_coverage_results_publisher} shows the coverage of
hybrid journals in transformative agreements by publisher portfolio
according to the ESAC registry.

>>>>

This study investigates the suitability of open scholarly data sources
for assessing the impact of transformative agreements on hybrid open
access. As shown in Figure \ref{fig:workflow}, the methodology involved
comparing hoaddata, an openly available collection of open research
information on hybrid open access, with the bibliometric databases Web
of Science and Scopus. This section introduces the initial data sources,
followed by a presentation of the necessary data processing steps to
obtain eligible articles from transformative agreements by using author
roles (first and corresponding) and harmonised affiliation data.

@@

As discussed elsewhere (Jahn, 2025), hoaddata uses a unified dataset of
historic JCT snapshots, which were safeguarded since July 2021. Journal
data were enriched with ISSN variants linked to an ISSN-L and insitution
data with associated ROR-IDs representing university hospitals or
institutes of large research organisations, according to OpenAlex'
institution entity. A comprehensive exclusion of fully open access
journals using multiple journal lists (DOAJ, OpenAlex, Bielefeld) was
performed to identify hybrid journals. hoaddata relies on Crossref for
obtaining journal publication volume and open access status through
Creative Commons licence information relative to the published version
(``version of record''). Because of limited affiliation metadata in
Crossref (Eck \& Waltman, 2022), hoaddata sources affiliation metadata
from OpenAlex.

hoaddata follows good practices for computational reproducibility using
R (Marwick et al., 2018). The package, which includes data, code, a test
suite and documentation, is openly available on GitHub. To ensure
computational reproducibility while aggregating the data, a GitHub
Actions continuous integration and delivery (CI/CD) workflow handles
data retrieval from the SUB Göttingen's open scholarly data warehouse
based on Google BigQuery, which provides high-performant programmatic
access to monthly snapshots of Crossref and OpenAlex. The workflow has
run regularly to fetch updates from these data sources since 2022. The
package version used in this study is 0.3, containing data from the
Crossref 2024-08 dump provided to Crossref Metadata Plus subscribers and
the OpenAlex 2024-08-29 monthly dump. This version including the
computation log is available on GitHub
(https://github.com/subugoe/hoaddata/releases/tag/v.0.3).

@@

Article metadata were retrieved from each database using all ISSN
variants linked to an ISSN-L. The metadata included DOIs, publication
dates, article types, open access information as well as author roles
and institutional affiliations. Publication years were determined using
the earliest known date of publication in a journal. In hoaddata, this
corresponded to Crossref's issued date. For Web of Science and Scopus,
the earliest publication date was used where available, with Scopus
dates specifically determined by the KB through version tracking of the
raw data.

@@

Published articles were considered hybrid open access when they were
made freely available under a Creative Commons license on publishers'
platforms. While hoaddata sourced this information from Crossref license
metadata, Web of Science and Scopus relied on Unpaywall. Unpaywall
supplements Crossref metadata by parsing publisher websites directly,
addressing cases where publishers do not provide machine-readable
Creative Commons license information (Piwowar et al., 2018). This
additional parsing remains necessary despite transformative agreement
workflows recommend the depositing license information during DOI
registration (Geschuhn \& Stone, 2017). Both Web of Science and Scopus
defined hybrid open access consistently as content available under
Creative Commons licenses on publisher platforms, distinguishing it from
bronze open access that lack such explicit license information.

@@

This process linked 6,375 ROR IDs to 4,894 Scopus Affiliation IDs, and
6,034 ROR IDs to 2,422 enhanced affiliation strings in the Web of
Science. Quality evaluation through random sampling of 50 pairs revealed
an error rate of 22\% for Web of Science (11 mismatches) and 6\% for
Scopus (3 mismatches). Upon inspection, these mismatches primarily
occurred with less-represented institutions having only a few
publications, introduced through multiple affiliations of single
authors. The difference between databases suggests that Scopus's
affiliation control aligns more closely with ROR than that of the Web of
Science.

@@

As a result of the comprehensive data processing described above,
datasets on open access in hybrid journals included in transformative
agreements were aggregated for each database at country and journal
level by year. Table~\ref{tbl-methods_overview_tab} provides a general
overview of the coverage between 2019 and 2023 per database. It shows
that the majority of hybrid journals published at least one original
research article or review marked as core during the five-year period.
These journals formed the basis for the subsequent calculation of
article-level indicators.

@@

\centering
\begin{tabular}[t]{lrrr}
\toprule
\textbf{} & \textbf{HOAD} & \textbf{Web of Science} & \textbf{Scopus}\\
\midrule
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Hybrid journal coverage}}\\
\hspace{1em}Active journals & 12,890 & 8,655 & 11,888\\
\hspace{1em}Active journals (core) & 12,888 & 8,655 & 11,878\\
\hspace{1em}Active journals (core) with OA & 11,348 & 8,392 & 11,313\\
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Publication volume}}\\
\hspace{1em}Total published articles & 9,740,015 & 8,616,053 & 8,117,644\\
\hspace{1em}Core articles & 8,158,425 & 6,708,083 & 7,317,703\\
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Digital Object Identifier (DOI) coverage}}\\
\hspace{1em}Articles with DOI & 9,740,015 & 7,713,796 & 8,105,112\\
\hspace{1em}Core articles with DOI & 8,158,425 & 6,695,661 & 7,314,327\\
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Open Access (OA) metrics}}\\
\hspace{1em}OA articles & 998,699 & 1,112,758 & 974,099\\
\hspace{1em}Core OA articles & 969,817 & 1,019,784 & 922,578\\
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Core articles with affiliation data}}\\
\hspace{1em}First author articles & 7,242,542 & 6,294,855 & 7,232,017\\
\hspace{1em}Corresponding author articles & 5,534,207 & 6,291,441 & 6,898,487\\
\bottomrule
\end{tabular}

@@

While hoaddata only covers articles with a DOI, Scopus and Web of
Science publication indicators were calculated using their database
identifier. A subsequent comparison of DOI coverage shows that non-core
articles in Web of Science often lacked a DOI. This was particularly the
case for meeting abstracts, which are notably prevalent in Health
Sciences journals (Melero-Fuentes et al., 2025) and are not indexed by
Scopus (Donner, 2017). Open access indicators were aggregated by DOI, as
Unpaywall only collects information on open access status for articles
with a DOI. A closer look at the core articles with affiliation data
reveals a lack of data on corresponding authors in the case of OpenAlex,
the affiliation data source used by hoaddata, compared to Web of Science
and Scopus. Only about two-thirds of the articles examined provided
corresponding author affiliations. For first authors, the proportion was
89\%. Therefore, only first author data for hoaddata were considered in
the following analysis.

@@

Figure \ref{fig-upset_coverage_results}B also shows DOIs that were only
available via hoaddata (grey area). In case of hybrid journals covered
by all three data sources, 1,023,882 DOIs were only present in hoaddata.
After validation at the DOI level using the KB databases and manual
inspection, the main reasons for missing DOI coverage in the proprietary
database were insufficient classification of journal content as original
research articles and reviews during the compilation of hoaddata.
Particulalry, letters and editorials could not be fully detected.
Paratext recognition failed in 37\% of DOIs to identify non-scientific
content such as front matters or reviewer lists, which are generally not
indexed by Scopus and Web of Science. To a lesser extent, differences in
publication and indexing dates were a reason for non-overlapping DOIs.

Using overlapping DOIs, the publication volume between 2019 and 2023 was
also calculated for each journal. Figure
\ref{fig-upset_coverage_results}C illustrates the distribution for each
combination. It shows that there is a large spread across the journals
covered by all three data sources. Furthermore, journals in this set
published more on average than journals covered by only one or two data
sources. In particular, the journals covered exclusively by hoaddata are
substantially smaller than those covered by all three sources. Upon
inspection, these are often newly launched journals, which explains the
relatively low five-year publication volume.

@@

Figure \ref{fig-upset_coverage_results_publisher} illustrates the
coverage of hybrid journals in transformative agreements by publisher
portfolio according to the ESAC registry between 2019 and 2023. Elsevier
was best represented in the union set of all investigated databases,
with 1,672 hybrid journals with at least one open access article that
published 2,345,205 articles between the five-year period. Springer
Nature followed with 1,561 hybrid journals publishing 1,058,564
articles, and Wiley with 1,151 hybrid journals and 770,546 articles. The
remaining 54 publishers recorded a total of 6,476 hybrid journals that
published 2,115,372 articles.

Notably, the three largest publishers--Elsevier, Springer Nature and
Wiley--were well represented in the exclusive intersection of all three
data sources examined. In terms of journal coverage, they accounted for
4,384 (55\%) of the hybrid journals included in the intersection. This
dominance is even more pronounced in terms of article coverage, with
these publishers having a joint indexing coverage of 4,174,315 articles
(66\%).
